---
title: "Modelling palaeoecological community data: a state-space approach"
author: "Quinn Asena, Anthony Ives, Jack Williams. and Jonathan Johnson"
institute: "UW-Madison"
date: today
bibliography: refs.bib
from: markdown+emoji
format:
  revealjs:
    theme: dark
#    backgroundcolor: "#44444c"
#    fontcolor: "white"
    preview-links: true
    highlight-style: github
    slide-number: c/t
    width: 1600
    height: 900
title-slide-attributes:
  data-background-color: "#1a1e43"
---


## Palaeoecological data

- Typically comprise mulitple time-series (response): 
  - e.g., pollen, diatoms...
- Include environmental covariates (predictors):
  - e.g., isotopic data, lake level...


Palaeoecological data pose many statistical challenges:

- Time averaging
- Uneven sampling through time
- Measurement uncertainty
- Relative abundances


## Descriptive approaches

Many descriptive approaches exist for analysing multivarivariate time-series:

- Cluster analyses: CONISS
- Ordination: PCA, NMDS...
- Machine learning: MVRT, LDA (fancy cluster analyses)

Descriptive methods allow us to establish patterns but not determine potential causes of those patterns.

## Beyond pattern recognision

The cutting edge in palaeoecology is to establish potential causes of observed patterns in species relative abundances. Are observed patterns driven by:

- species interactions?
- climate variability?
- fire regime?

No easy task!


## State-space modelling

State-space modelling attempts to predict the "true" unobservable state of a system from observable variables. It does so via two equations, one that models the _process of the system_:

<br>
Process equation:
$$
z_t = z_{t-1} + \epsilon_t,   \epsilon_t \sim N(0, \sigma^2_p)
$$

and one that models the _observations_ from the system:

<br>

Observation equation:
$$
y_t = z_y + n_t,   n_t \sim N(0, \sigma^2_o)
$$

## State-space modelling

State-space modeling goes beyond descriptive approaches and attempts to estimate:

- autoregressive processes
- species interactions
- species-environment interactione
- combinations of the above

Things get complicated when esimating multiple parameters simultaneously!

## State-space modelling

This variant of a state-space model:

- uses a multinomial distribution: accepts raw count data
- accepts multiple predictor data streams in the same model: e.g., isotopic data, charcoal accumulation rates...
- simultaneously fits multiple coefficients
- models autocorrelation structure
- assess a range of possible causes of observed patterns


## Empirical example

Demonstrating a three-species model from Sunfish Pond:

- unpublished data: not presenting the dataset, focusing on the modelling approach
- In this example, the explanitory variable is time. I.e., we are looking at change in taxa over time. 
- interested in abrupt transitions between dominant species

```{r map}
library(leaflet)
m <- leaflet() %>%
  setView(lng = -76.6990733, lat = 41.6433326, zoom = 5) %>%
  addMarkers(lng = -76.6990733, lat = 41.6433326, popup="Sunfish Pond")
m %>% addTiles()
```

## Empirical example

```{r, plot}
library(patchwork)
two_spp_plt <- readRDS("./images/two_spp_plt.rds")
two_spp_plt

```


## Simulation evaluation

- We cannot determine the accuracy of fitted coefficients empirically from palaeoecological data. Experimental manipulation is challenging.

- Simulation experiments are used to assess the success of recovering parameters:
  1. data are simulated under known conditions with replication
  2. the model is fit to the simulated replicate datasets
  3. fitted models are assessed for how well input parameters are recovered

## Simulation evaluation

Example of recovering parameters for inter- intra- specific interactions:

## Hypothesis testing

We cannot determine with certainty, outside of simulation, causation from palaeo-data.

What we can do is:
- set up multiple working hypothese (Chamberlin)
- couple descriptive methods with inferrential ones
- assess which results lend support to the likelihood of each hypothesis being true



## Acknowledgements

